---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cytotypr

<!-- badges: start -->
<!-- badges: end -->

The goal of cytotypr is to ...

## Installation

You can install the development version of cytotypr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("aef1004/cytotypr")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(cytotypr)
## basic example code
```





## Full Example
Note that I currently have only downloaded a few flow cytometry files. They are from Cyto feature engineering T cell panel.

```{r message = FALSE}
library(data.table)
library(openCyto)
library(ncdfFlow)
library(flowWorkspace)
library(dplyr)
library(ggcyto)
library(stringr) #this should be already added to the cytotypr package
library(scales)
library(tidyr)
library(superheat)
```


Read in the initial gating strategy, which gates to live lymphocytes, into a `gatingTemplate` class.

```{r}
# Identify the file with the gating strategy 
ws <- list.files("./inst/extdata/", 
                 pattern = "gating_strategy.csv", 
                 full = TRUE)
ws

```

```{r}
# View this template
dtTemplate <- fread(ws)
dtTemplate
```

```{r message = FALSE, warning = FALSE}
# Read in the gating strategy to a 'gatingTemplate' object
initial_gate <- gatingTemplate(ws) 
plot(initial_gate)
```
### Read in the FMO data
```{r}
# Identify the file names of all 20 FCS flow cytometry experiments to 
# read in. 
FMO_fcsFiles <- list.files("./inst/extdata/FMOs", full = TRUE)
FMO_fcsFiles

```

```{r}
# Read these files into a 'ncdfFlowSet' object. This will taken a minute
# or so to run. The resulting 'ncdfFlowset' object contains row names with the 
# individual samples and column names with the markers/parameters used in the flow cytometer.
ncfs_FMO <- read.ncdfFlowSet(FMO_fcsFiles, channel_alias = data.frame(alias = c("Zombie Nir-A"), channels = c("Zombie_NIR-A, Zombie NIR-A, Zombie Nir-A"))) 
ncfs_FMO
```

Apply the initial gating to filter the data to only measurements on live lymphocyte cells.

```{r}
# Convert to a 'GatingSet' object, so we can apply the initial gating to this data.
gs_FMO <- GatingSet(ncfs_FMO)
gs_FMO
```

```{r fig.height = 8, fig.width = 4}
# Apply the initial gating to this data, to filter to only measurements
# on live lymphocyte cells. This may take a minute.
gt_gating(initial_gate, gs_FMO)

# You can plot the results of this gating with `plotGate`. For example, to plot
# the gating for the first sample, run:
plotGate(gs_FMO[[1]])
```

## Convert data to "tidy data" format

Now that the initial gating has been applied, to limit the data to measurements oflive, singlet lymphocyte cells, we convert the data to a "tidy data" format, to allow us to work with "tidyverse" tools for further analysis and visualization.

```{r}
# Pull out the data from the 'live' node of the gating set (the last node
# in the initial gating strategy).
flowset_FMO_gated_data <- gs_pop_get_data(gs_FMO, "live") %>% 
  as.flowSet() 
```

Apply this function to the 'flowSet' of gated FMO data:

```{r}
FMO_gated_data <- tidy_flow_set(flowset_FMO_gated_data)
FMO_gated_data
```

```{r}
# note that here the filename and the column marker names need to match exactly
df_FMO_gated_data <- FMO_gated_data %>%
  select(ends_with("-A"), -`FSC-A`, `SSC-A`, filename) %>%
  dplyr::rename(`FoxP3` = "APC-A",
         `CD44` = "APC-Fire 750-A",
        `CD103` =  "APC-R700-A",       
         `CD3` = "Alexa Fluor 532-A",
         `Sca1` = "BB515-A",
         `IL_10` = "BV421-A",
         `CD4` = "BV480-A",
         `CD69` = "BV510-A",
         `CD8` = "BV570-A", 
         `CTLA4` = "BV605-A",
         `CD27` = "BV650-A",
         `CD153` = "BV711-A",
         `KLRG1` = "BV785-A",
         `IL_17` = "PE-A",
         `CD122` = "PE-Cy5-A",
         `IFN` = "PE-Cy7-A", 
         `CD62L` = "PE-Dazzle594-A",
         `TNF` = "Pacific Blue-A", 
         `CD28` = "PE-Cy5.5-A",
         `PD1` = "PerCP-eFluor 710-A")  %>%
   na.omit()%>%
  dplyr::filter(`SSC-A` != max(`SSC-A`)) %>%
  mutate(filename = str_replace(filename, ".fcs", "")) %>%
  mutate(filename = str_replace(filename, "IFNG", "IFN"))

  
FMO_filtered_data <- filter_FMO(df_FMO_gated_data)

add_quantile <- get_99(FMO_filtered_data)

plot_FMOs(FMO_filtered_data, add_quantile)
```


### Gating the sample data

I checked that all D30, D60, and D90 samples have Zombie and will run all the way through the gating when separated by day. Looking at the data, some of them are labeled Zombie NIR-A and Zombie Nir-A which means that they are not recognized as the same. - D90 has uncapitalized. I will need to read them in separately and then rename one of the two so that the cases match

```{r warning=FALSE, fig.height=1.5, fig.width=1.5}

fcsFiles <- list.files("./inst/extdata/Tcell_samples", 
                       pattern = ".fcs", full = TRUE)

# ncdfFlowset object contains row names with the individual samples and column names with the markers/parameters used in the flow cytometer
ncfs <- read.ncdfFlowSet(fcsFiles, channel_alias = data.frame(alias = c("Zombie Nir-A"), channels = c("Zombie_NIR-A, Zombie NIR-A, Zombie Nir-A"))) 

# apply gating set
gs <- GatingSet(ncfs)

# gate the samples
gt_gating(initial_gate, gs)

# View the gates
plotGate(gs[[1]])

ggcyto(gs[[3]], aes(x = `Zombie Nir-A`)) + 
  geom_area(stat = "density", fill = "#440154FF") +
  geom_gate("live")   +
  ggcyto_par_set(limits = list(y = c(0, 5e-5), x = c(-1000, 1e5)))+
  xlab("Zombie NIR-A") +
  ylab("Density") +
  ggtitle("Live") +
  theme_gray() +
  geom_stats(size = 5, adjust = 0.45, position = position_nudge(x = 5000)) +
  labs_cyto("both") + 
  theme(strip.text = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 20),
        title = element_text(size = 20))+
  scale_x_continuous(n.breaks = 3)

```

Pull out the gated data
```{r}
# Pull out the gated data - could potentially add to the function below
gated_flowset <- gs_pop_get_data(gs, "live") %>% 
  as.flowSet() 

# tidy the flowset and convert to a dataframe
df_all_gated <-  tidy_flow_set(gated_flowset) 

unique(df_all_gated$filename)
```

Feature engineer the data

Feature cut the data to get all of the possible populations for each file with the cell count and number of cells. Feature engineering is based on the 99%.

Remove FoxP3
Remove CD69 - spreading from Sca1 makes the marker unusable

```{r}
all_fe <- df_all_gated %>%
  mutate(Timepoint = str_extract(filename, "D[0-9]*")) %>%
  mutate(Group = str_extract(filename, "Group[:punct:][0-9][:punct:][A-Z]"),
         Group = str_replace(Group, "Group_", "")) %>%
  unite(filename, c("Timepoint", "Group")) %>%
  select(ends_with("-A"), -`FSC-A`, filename) %>%
  dplyr::rename(`FoxP3` = "APC-A",
         `CD44` = "APC-Fire 750-A",
        `CD103` =  "APC-R700-A",       
         `CD3` = "Alexa Fluor 532-A",
         `Sca1` = "BB515-A",
         `IL_10` = "BV421-A",
         `CD4` = "BV480-A",
         `CD69` = "BV510-A",
         `CD8` = "BV570-A", 
         `CTLA4` = "BV605-A",
         `CD27` = "BV650-A",
         `CD153` = "BV711-A",
         `KLRG1` = "BV785-A",
         `IL_17` = "PE-A",
         `CD122` = "PE-Cy5-A",
         `IFN` = "PE-Cy7-A", 
         `CD62L` = "PE-Dazzle594-A",
         `TNF` = "Pacific Blue-A", 
         `CD28` = "PE-Cy5.5-A",
         `PD1` = "PerCP-eFluor 710-A") %>%
  dplyr::filter(`SSC-A` != max(`SSC-A`)) %>%
  mutate(CD3 = fe(add_quantile, CD3, "CD3"),
         CD4 = fe(add_quantile, CD4, "CD4"),
         CD8 = fe(add_quantile, CD8, "CD8"),
         CD44 = fe(add_quantile, CD44, "CD44"),
         CD103 = fe(add_quantile, CD103, "CD103"),
         Sca1 = fe(add_quantile, Sca1, "Sca1"),
         IL_17 = fe(add_quantile,IL_17, "IL_17"),
         CD69 = fe(add_quantile,CD69, "CD69"),
         CTLA4 = fe(add_quantile,CTLA4, "CTLA4"),
         CD27 = fe(add_quantile,CD27, "CD27"),
         CD153 = fe(add_quantile,CD153, "CD153"),
         KLRG1 = fe(add_quantile,KLRG1, "KLRG1"),
         IFN = fe(add_quantile,IFN, "IFN"),
         FoxP3 = fe(add_quantile,FoxP3, "FoxP3"),
         CD122 = fe(add_quantile,CD122, "CD122"),
         PD1 = fe(add_quantile,PD1, "PD1"),
         CD62L = fe(add_quantile,CD62L, "CD62L"),
         IL_10 = fe(add_quantile,IL_10, "IL_10"),
         CD28 = fe(add_quantile,CD28, "CD28"),
         TNF = fe(add_quantile,TNF, "TNF")) %>%
  select(-`Zombie Nir-A`, -`AF-A`, -`SSC-A`, -FoxP3, -CD69) %>%
  count_calc()

```

Visulatizations

Initial identification of populations plot

We first want to view all of the different cell phenotypes within the data

```{r}
# this is the order of markers that we want for all of our plots
order_of_markers <- c("CD3", "CD4", "CD8",  "CD44", "CD103", "Sca1", "IL_17","CTLA4",
                      "CD27",  "CD153", "KLRG1", "IFN",  "CD122", "PD1", "CD62L",
                      "IL_10", "CD28","TNF")

# to view all of the possible combinations
total_phenotypes <- filter_for_total_pheno(all_fe, marker_order = order_of_markers)

heatmap_all_pheno(total_phenotypes)

# gives the total number of populations
nrow(total_phenotypes) 
```


After identifying all phenotypes, we can filter the data to see the ones that we're interested in, for example, CD3+ cells that constitute >0.5% of total live leukocytes in a sample.

```{r}
# view the specific cell phenotypes we're interested in
sample_populations <- all_fe %>%
  dplyr::filter(CD3 == 1 & percentage > 0.5) %>%
  filter_pops() 

sample_populations_all_groups <- identified_pop_perc(sample_populations, all_fe, marker_vector = order_of_markers)
```

Plot sample populations
```{r}

```


Correlation plot

```{r}

corr <- calc_corr(sample_populations_all_groups)

# superheat(corr,  row.title = "Populations", column.title = "Populations",  column.title.size = 8, row.title.size = 8, bottom.label.col = c(rep("darkturquoise", DN_length), rep("plum3", T_helper_length), rep("hotpink", Cytotoxic_length), rep("orange", DP_length)), left.label.col = c(rep("darkturquoise", DN_length), rep("plum3", T_helper_length), rep("hotpink", Cytotoxic_length), rep("orange", DP_length)), bottom.label.size = .05, left.label.size = .05, bottom.label.text.size = 5, left.label.text.size = 5)

```

